---
title: Manipulating, analyzing and exporting data with the tidyverse
author: Cecilia Noecker & Data Carpentry contributors
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Describe the purpose of the **`dplyr`** and **`tidyr`** packages.
> * Select certain columns in a data frame with the **`dplyr`** function `select`.
> * Select certain rows in a data frame according to filtering conditions with the **`dplyr`** function `filter` .
> * Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`.
> * Add new columns to a data frame that are functions of existing columns with `mutate`.
> * Use the split-apply-combine concept for data analysis.
> * Use `summarize`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.
> * Describe the concept of a wide and a long table format and for which purpose those formats are useful.
> * Reshape a data frame from long to wide format and back with the `pivot_wider` and `pivot_longer` commands from the **`tidyr`** package.
> * Export a data frame to a .csv file.

------------

# Data Manipulation using **`dplyr`** and **`tidyr`**

How much time do computational biologists spend on "data wrangling"? Answer: A Lot. The tidyverse is an amazing set of tools for making these tasks easier and more streamlined.

Bracket subsetting is handy, but it can be cumbersome and difficult to read,
especially for complicated operations. Enter **`dplyr`**. **`dplyr`** is a package for
making tabular data manipulation easier. It pairs nicely with **`tidyr`** which enables you to swiftly convert between different data formats for plotting and analysis.

Packages in R are basically sets of additional functions that let you do more
stuff. The functions we've been using so far, like `str()` or `data.frame()`,
come built into R; packages give you access to more of them. Before you use a
package for the first time you need to install it on your machine, and then you
should import it in every subsequent R session when you need it. You should
already have installed the **`tidyverse`** package. This is an
"umbrella-package" that installs several packages useful for data analysis which
work together well such as **`tidyr`**, **`dplyr`**, **`ggplot2`**, **`tibble`**, etc.


The **`tidyverse`** package tries to address 3 common issues that arise when
doing data analysis with some of the functions that come with R:

1. The results from a base R function sometimes depend on the type of data.
2. Using R expressions in a non standard way, which can be confusing for new
   learners.
3. Hidden arguments, having default operations that new learners are not aware
   of.

We have seen in our previous lesson that when building or importing a data frame, the columns that contain characters (i.e., text) are coerced (=converted) into the `factor` data type. We had to set **`stringsAsFactors`** to **`FALSE`** to avoid this hidden argument to convert our data type. 

This time we will use the **`tidyverse`** package to read the data and avoid having to set **`stringsAsFactors`** to **`FALSE`**

If we haven't already done so, we can type `install.packages("tidyverse")` straight into the console. In fact, it's better to write this in the console than in our script for any package, as there's no need to re-install packages every time we run the script.

Then, to load the package type:


```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

## What are **`dplyr`** and **`tidyr`**?

The package **`dplyr`** provides easy tools for the most common data manipulation
tasks. It is built to work directly with data frames, with many common tasks
optimized by being written in a compiled language (C++). An additional feature is the
ability to work directly with data stored in an external database. 

This addresses a common problem with R in that all operations are conducted
in-memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can connect to a database of many hundreds of GB, conduct queries on it directly, and pull
back into R only what you need for analysis.

The package **`tidyr`** addresses the common problem of wanting to reshape your data for plotting and use by different R functions. Sometimes we want data sets where we have one row per measurement. Sometimes we want a data frame where each measurement type has its own column, and rows are instead more aggregated groups - like plots or aquaria. Moving back and forth between these formats is non-trivial, and **`tidyr`** gives you tools for this and more sophisticated  data manipulation.

To learn more about **`dplyr`** and **`tidyr`** after the workshop, you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) and this [one about **`tidyr`**](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf).

We'll read in our data using the `read_delim()` function, from the tidyverse package **`readr`**, instead of `read.csv()`.


```{r, results = 'hide', purl = FALSE}
metadata <- read_delim("~/Desktop/tutorial16S/metadata.tsv", delim = "\t")
metadata <- read_delim("~/Desktop/tutorial16S/metadata.tsv", delim = "\t", comment = "#")
```

You will see the message `Parsed with column specification`, followed by each column name and its data type.
When you execute `read_delim` on a data file, it looks through the first 1000 rows of each column and
guesses the data type for each column as it reads it into R. For example, in this dataset, `read_delim`
reads `weight` as `col_double` (a numeric data type), and `species` as `col_character`. You have the
option to specify the data type for a column manually by using the `col_types` argument in `read_delim`.

```{r, results = 'hide', purl = FALSE}
## inspect the data
str(metadata)
```

```{r, eval=FALSE, purl=FALSE}
## preview the data
View(metadata)
```

Notice that the class of the data is now `tbl_df`

This is referred to as a "tibble". Tibbles tweak some of the behaviors of the data frame objects we introduced in the previous episode. The data structure is very similar to a data frame. For our purposes the only differences are that:

1. In addition to displaying the data type of each column under its name, it
   only prints the first few rows of data and only as many columns as fit on one
   screen.
2. Columns of class `character` are never converted into factors.


We're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values
- `[]_join()`: merge two data frames

## Selecting columns and filtering rows

To select columns of a data frame, use `select()`. The first argument
to this function is the data frame (`metadata`), and the subsequent
arguments are the columns to keep.

```{r, results = 'hide', purl = FALSE}
select(metadata, MouseID, Treatment)
```

To select all columns *except* certain ones, put a "-" in front of
the variable to exclude it.

```{r, results = 'hide', purl = FALSE}
select(metadata, -SampleID)
```

There are also fancier ways to select columns, including the function `starts_with()`: 

```{r, purl=FALSE}
select(metadata, starts_with("T"))
```


To choose rows based on a specific criterion, use `filter()`:

```{r, purl = FALSE}
filter(metadata, Treatment == "After_Diet")
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame and use
that as input to the next function, like this:

```{r, purl = FALSE}
metadata2 <- filter(metadata, Time_days < 7)
metadata_sml <- select(metadata2, SampleID, Time_days)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl = FALSE}
metadata_sml <- select(filter(metadata, Time_days < 7), SampleID, Time_days)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take
the output of one function and send it directly to the next, which is useful
when you need to do many things to the same dataset.  Pipes in R look like
`%>%` and are made available via the **`magrittr`** package, installed automatically
with **`dplyr`**. If you use RStudio, you can type the pipe with <kbd>Ctrl</kbd>
+ <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> + 
<kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl = FALSE}
metadata %>%
  filter(Time_days < 7) %>%
  select(SampleID, Time_days)
```

In the above code, we use the pipe to send the `metadata` dataset first through
`filter()` to keep rows where `weight` is less than 5, then through `select()`
to keep only the `species_id`, `sex`, and `weight` columns. Since `%>%` takes
the object on its left and passes it as the first argument to the function on
its right, we don't need to explicitly include the data frame as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we took the data frame `metadata`, *then* we `filter`ed
for rows with `weight < 5`, *then* we `select`ed columns `species_id`, `sex`,
and `weight`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl = FALSE}
metadata_sml <- metadata %>%
  filter(Time_days < 7) %>%
  select(SampleID, Time_days)

metadata_sml
```

Note that the final data frame is the leftmost part of this expression.

> ### Challenge {.challenge}
>
>  Using pipes, subset the `metadata` data to include samples before diet only, and retain only the columns 
> `MouseID` and `Time_days`.
> 
> 
> ```{r, answer=TRUE, eval=FALSE, purl=FALSE}
> metadata %>%
>     filter(Treatment == "Before_Diet") %>%
>     select(MouseID, Time_days)
> ```

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
##  Using pipes, subset the `metadata` data to include samples before diet only, and retain only the columns 
## `MouseID` and `Time_days`.
```

### Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column with Time and Treatment combined:

```{r, purl = FALSE}
metadata %>%
  mutate(Group = paste(Treatment, Time_days, sep = "_"))
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl = FALSE}
metadata %>%
  mutate(Group = paste(Treatment, Time_days, sep = "_"),
         GroupShort = gsub("Diet_", "", Group))
```

If this runs off your screen and you just want to see the first few rows, you
can use a pipe to view the `head()` of the data. (Pipes work with non-**`dplyr`**
functions, too, as long as the **`dplyr`** or `magrittr` package is loaded).

```{r, purl = FALSE}
metadata %>%
  mutate(Group = paste(Treatment, Time_days, sep = "_"),
         GroupShort = gsub("Diet_", "", Group)) %>% head()
```

We can also combine mutate with filter, select, and any other similar function. 

```{r, purl = FALSE}
metadata %>%
  filter(!is.na(Time_days)) %>%
  mutate(Group = paste(Treatment, Time_days, sep = "_")) %>%
  head()
```

`is.na()` is a function that determines whether something is an `NA`. The `!`
symbol negates the result, so we're asking for every row where weight *is not* an `NA`.

Let's read in the feature table to test out these tools. To do so, we will use the **`read_qza`** function in the **`qiime2R`** package.

```{r, purl = FALSE}
library(qiime2R)
feature_table <- read_qza("~/Desktop/tutorial16S/artifacts/Feature_Table.qza")$data
class(feature_table)

```

The feature table is imported as a matrix. We'll convert it to a tibble so that we can manipulate it more easily. We'll also convert the row names (sequence IDs) into a column of the tibble, named FeatureID.

```{r, purl = FALSE}
feature_table <- feature_table %>% data.frame() %>% rownames_to_column(var = "FeatureID") %>% as_tibble()
class(feature_table)
names(feature_table)
str(feature_table)

```



> ### Challenge {.challenge}
>
>  Select the FeatureID column and one other sample column from the feature table, filter it to rows where its value is greater than 100 reads, and make a new column representing the relative abundance of the remaining reads. 

> ```{r, answer=TRUE, eval=FALSE, purl=FALSE}
> feature_sub <- feature_table %>%
>     select(FeatureID, MOUSE.H13.Day3) %>% 
>     filter(MOUSE.H13.Day3 > 100) %>%
>     mutate(H13_3_rel = MOUSE.H13.Day3 / sum(MOUSE.H13.Day3)
>     
> ```
 

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Mutate Challenge:
## Select the FeatureID column and one other sample column from the feature table, filter it to rows where
## its value is greater than 100 reads, and make a new column representing the relative abundance of the
## remaining reads.

```

## Reshaping data with `pivot_longer()` and `pivot_wider()`

You can imagine it would be annoying to repeat the same commands if we want to calculate relative abundances for all samples individually. Our current feature table is set up to compare how each feature ID (rows) varies across all the different samples. Instead, we want to reorganize the feature table so each observation to have its own row, which will allow us to compare them across Samples, Features, and other variables. We can do this transformation, and its reverse, using the `tidyr` package. Previously, the recommended functions to do this were called "spread" and "gather", which you can still use if you've used them before, but more intuitive functions to do the same thing have recently been introduced: `pivot_longer()` and `pivot_wider()`.

Here, we want to make our feature table longer (more rows, fewer columns).

#### Pivot longer

`pivot_longer()` takes four main arguments:

1. the data 
2. the names of columns we want to melt into a single column.  
3. the name of the new column of column names.
4. the name of the new column of values.



```{r, purl=FALSE}
features_long <- pivot_longer(feature_table, -FeatureID, names_to = "SampleID", values_to = "Reads")

dim(features_long)
dim(feature_table)
head(features_long)
```


#### pivot_wider()

`pivot_wider()` performs the reverse operation: 

```{r, purl=FALSE}
features_rev <- pivot_wider(features_long, names_from = "SampleID", values_from = "Reads")

```

## Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of the
`group_by()` function.


#### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. Let's try out a few options to summarize the average number of reads for each feature and see those with the most and least reads: 

```{r, purl = FALSE}
features_long %>%
  group_by(FeatureID) %>%
  summarize(AvgReads = mean(Reads, na.rm = TRUE))


features_long %>%
  group_by(FeatureID) %>%
  summarize(AvgReads = mean(Reads, na.rm = TRUE)) %>% 
  arrange(AvgReads) %>% 
  print(n = 15)

features_long %>%
  group_by(FeatureID) %>%
  summarize(AvgReads = mean(Reads, na.rm = TRUE)) %>% 
  arrange(AvgReads) %>% 
  filter(AvgReads < 1)

features_long %>%
  group_by(FeatureID) %>%
  summarize(AvgReads = mean(Reads, na.rm = TRUE)) %>% 
  arrange(desc(AvgReads)) %>% 
  print(n = 15)

```

Why is it important to calculate these kinds of exploratory statistics? What else might we want to check?

Let's add another column for relative abundance so this is more informative.

```{r, purl = F}
features_long %>% 
  group_by(SampleID) %>% 
  mutate(RelAbund = Reads/sum(Reads))

features_long <- features_long %>% 
  group_by(SampleID) %>% 
  mutate(RelAbund = Reads/sum(Reads)) %>% 
  ungroup()

```

What if we want to summarize feature abundances across a metadata variable like Treatment group or day? We need to join the two tables together. There are a set of dplyr functions to do this: `inner_join()`, `full_join()` , `left_join()`, and `right_join()`. The type of join that you want to do depends on whether you want to keep just the shared observations between your two tables, or all observations from one or both tables. You can read more about join functions by typing `?join`. 

```{r, purl = FALSE}
features_long <- left_join(features_long, metadata, by = "SampleID")

```

Some other convenient summary functions: 

```{r, purl = FALSE}
features_long %>%
  filter(Reads > 0) %>% 
    count(Treatment) 
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `surveys %>% count()` is equivalent to:  

```{r, purl = FALSE}
features_long %>%
    filter(Reads > 0) %>% 
    group_by(Treatment) %>%
    summarize(count = n())
```

For convenience, `count()` provides the `sort` argument:  

```{r, purl = FALSE}
features_long %>%
    filter(Reads > 0) %>% 
    count(SampleID, sort = T)
```

Try calculating some other summary statistics: 

- How many reads total in each sample?
- How many features have nonzero abundances in each mouse?
- What are the 10 most abundant features in each treatment group?
- How many features have < 5 reads in all samples? Would you want to remove these for downstream analysis?
- Which features have the biggest difference in average abundance between treatment groups?

# Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_delim()` function used for reading data files into R, there is
a `write_csv()` function that generates CSV files from data frames. Let's make a reduced table of only ASVs with at least 1% relative abundance in at least 1 sample, and save it to a file.


```{r, purl=FALSE}
features_keep <- features_long %>% 
  filter(RelAbund > 0.01) %>% 
  count(FeatureID)

features_keep <- inner_join(features_long, features_keep, by = "FeatureID") %>% 
  select(FeatureID, SampleID, MouseID, Treatment, Time_days, Reads, RelAbund)


write_csv(features_keep, path = "data/FeatureTableFiltered.csv")
```

> ### Challenge {.challenge}
>
>  Import the taxonomy table, and join it to the ASV feature table. You will need to first read it in as a qza 
>  object, convert it to a tibble, and then join it. Hint: there is a `parse_taxonomy()` function in qiime2R that 
>  should be helpful.
> 
> 
> ```{r, answer=TRUE, eval=FALSE, purl=FALSE}
> taxonomy <- read_qza("artifacts/Feature_Taxonomy.qza")$data %>% 
>     parse_taxonomy(tax_sep = "; ") %>% 
>     rownames_to_column(var = "FeatureID") %>% 
>     as_tibble()
>
> features_tax <- left_join(features_keep, taxonomy, by = "FeatureID")
>
> write_csv(features_tax, path = "data/FeaturesFilteredTaxonomy.csv")
> ```


```{r, child="_page_built_on.Rmd"}
```
